---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(tidyverse)
library(DiagrammeR)
library(knitr)
library(igraph)
devtools::load_all()
library(DemoKin)
```

# DemoKin

`DemoKin` uses matrix demographic methods to compute expected (average) kin counts from demographic rates under a range of scenarios and assumptions. The package is an R-language implementation of Caswell (2019) and Caswell and Song (2021). It draws on previous theoretical development by Goodman, Keyfitz and Pullum (1974).

## Installation

You can install the development version from GitHub with:

``` {r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("IvanWilli/DemoKin")
```

## Example

Consider an average Swedish woman in year 2015 (i.e., 'Focal'). 
Relatives for the `selected_kin` argument are identified by a unique code.
Note that the relationship codes used in `DemoKin` differ from those in Caswell (2019). 
The equivalence between the two set of codes is given in the following table:

```{r, fig.height=6, fig.width=8, echo=FALSE}

codes <- c("coa", "cya", "d", "gd", "ggd", "ggm", "gm", "m", "nos", "nys", "oa", "ya", "os", "ys")
caswell_codes <- c("t", "v", "a", "b", "c", "h", "g", "d", "p", "q", "r", "s", "m", "n")
labels <- c("Cousins from older aunt", "Cousins from younger aunt", "Daughter", "Grand-daughter", "Great-grand-daughter", "Great-grandmother", "Grandmother", "Mother", "Nieces from older sister", "Nieces from younger sister", "Aunt older than mother", "Aunt younger than mother", "Older sister", "Younger sister")

data.frame(DemoKin = codes, Caswell = caswell_codes, Label = labels, row.names = NULL) %>% kable(caption = "Equivalence between relative codes between DemoKin and Caswell (2019).")

```

Let's show a quick example. 
We assume a female closed population in which everyone experiences the Swedish 2015 mortality and fertility rates at each age throughout their life.
We then ask:

> How can we characterize the kinship network of an average member of the population ('Focal')?

For this exercise, we'll use the Swedish data pre-loaded with `DemoKin`.

```{r, fig.height=6, fig.width=8}
library(DemoKin)
swe_surv_2015 <- swe_px[,"2015"]
swe_asfr_2015 <- swe_asfr[,"2015"]
swe_2015 <- kin(U = swe_surv_2015, f = swe_asfr_2015, time_invariant = TRUE)
```

*px* is the survival probability by age from a life table and *f* are the age specific fertility ratios by age (see `?kin` for details). 

For more details, including an extension to time varying populations rates and relative´s death distribution, see `vignette("Use")`.
Note that if the vignette does not load, you may need to install the package as `devtools::install_github("IvanWilli/DemoKin", build_vignettes = T)`.

## Citation

Williams, Iván; Alburez-Gutierrez, Diego; Song, Xi; and Hal Caswell. (2021) DemoKin: An R package to estimate kinship networks in stable and non-stable populations. URL: https://github.com/IvanWilli/DemoKin.

## References

Caswell, H. (2019). The formal demography of kinship: A matrix formulation. Demographic Research 41:679–712. doi:10.4054/DemRes.2019.41.24.

Caswell, Hal and Xi Song. 2021. “The Formal Demography of Kinship. III. Kinship Dynamics with Time-Varying Demographic Rates.” Demographic Research 45: 517–46. doi:10.4054/DemRes.2021.45.16.

Goodman, L.A., Keyfitz, N., and Pullum, T.W. (1974). Family formation and the frequency of various kinship relationships. Theoretical Population Biology 5(1):1–27. doi:10.1016/0040-5809(74)90049-5.

## Next steps:

1. Implement multi-stage and two-sex matrix kinship models
1. Create a hex logo for package
1. Improve kinship diagram visualization
1. Improve documentation and vignette of package 

## Get involved!
  
`DemoKin` is giving its first steps.
If you're interested in contributing, please get in touch, create an issue, or submit a pull request.
We look forward to hearing from you!
