---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(tidyverse)
library(DiagrammeR)
library(knitr)
devtools::load_all()
```

# DemoKin

This package uses matrix demographic methods to estimate kin counts and the age distribution of relatives in stable and non-stable populations.
The package is an implementation of Caswell (2019) and draws on previous theoretical development by Goodman, Keyfitz and Pullum (1974).
`DemoKin` is giving its first steps, so please contact us or submit a pull request if you have any suggestion.

## Installation

You can install the development version from GitHub with:

``` {r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("IvanWilli/DemoKin")
```

## Example

``` {r,eval=F}
library(DemoKin)

library(tidyverse)
library(DiagrammeR)
library(knitr)
```


Consider an average Swedish woman aged 50 in 2015.
How many relatives can this woman expect to have?
The function `kins` can be used to obtain the expected number of Ego's relatives for the following types of relative:

```{r, fig.height=6, fig.width=8, echo=FALSE}

relatives <- c(
  "coa" = "Cousins (through aunt older than mother)"
  , "cya" = "Cousins (through aunt younger than mother)"
  , "d" = "Daughter"
  , "gd" = "Grand-daughter"
  , "ggm" = "Great-grandmother"
  , "gm" = "Grandmother"
  , "m" = "Mother"
  , "nos" = "Nieces through older sister"
  , "nys" = "Nieces through younger sister"
  , "oa" = "Aunt older than mother"
  , "ya" = "Aunt younger than mother"
  , "os" = "Older sister"
  , "ys" = "Younger sister"
  )

data.frame(Code = names(relatives), Relative = relatives, row.names = NULL) %>% kable()

```

For this example, we assume demographic stability (i.e., we assume that the womans and her relatives experienced the mortality and fertility rates from 2015 at each age throughout their life):

```{r, fig.height=8, fig.width=10}
swe50_2015_stable <- kins(ego_age = 50, year = 2015,
                             P = swe_surv, asfr = swe_asfr,
                             stable = TRUE)

plot_diagramm(swe50_2015_stable[["kins_total"]],ego_age = 50)
```

For more details, including an extension to non-stable populations, see `vignette("Use")`.

## Citation

Williams, Ivan and Diego Alburez-Gutierrez. (2021) DemoKin: An R package to implement matrix kinship models in stable and non-stable populations. URL: https://github.com/IvanWilli/DemoKin.

## References

Caswell, H. (2019). The formal demography of kinship: A matrix formulation. Demographic Research 41:679–712. doi:10.4054/DemRes.2019.41.24.

Caswell, H. (2020). The formal demography of kinship II: Multistate models, parity, and sibship. Demographic Research 42:1097–1146. doi:10.4054/DemRes.2020.42.38.

Goodman, L.A., Keyfitz, N., and Pullum, T.W. (1974). Family formation and the frequency of various kinship relationships. Theoretical Population Biology 5(1):1–27. doi:10.1016/0040-5809(74)90049-5.


## Next steps:

- Improve performance of `kins_non_stable` function.
- Give an option to forecast mortality and fertility to locate Ego and relatives in the future.
- Add more functionalities to the diagram, like colors by kin degree and box size weighted by kin amount.
- Add cumulative deaths by kin in the non-stable case, and also stage properties as in Caswell (2020). 
