---
title: "Technical details"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
    theme: readable
    highlight: pygments
    df_print: paged
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
pkgload::load_all()
```

## Location of deaths

In the time-varying models, **living** kin are *stocks*: they refer to the expected number of relatives alive at the beginning of the period, by the Focal's age, the kin's age, and calendar year. **Deaths** are *flows* over the *following year of risk*. In the output, those deaths are located at the year and Focal age at which the risk interval begins.

This means that, for a given kin type and kin age, the ratio between deaths during the year and living kin at the start of the year should reproduce the period death probability, `1 - p`. The small example below uses the one-sex time-varying model and checks this relationship for mothers.

```{r death-location-example}

years_use <- c("1900", "1901", "1902")

# model
kin_mothers <- kin_time_variant(
  p = swe_px[, years_use],
  f = swe_asfr[, years_use],
  output_period = as.integer(years_use[1:2]),
  output_kin = "m"
)

# pick a kin to check
check_mothers <- subset(
  kin_mothers,
  year == 1900 &
    kin == "m" &
    age_focal == 0 &
    living > 0
)

# check probabilities
check_mothers$death_over_living <- check_mothers$dead / check_mothers$living
check_mothers$period_death_prob <- 1 - swe_px[check_mothers$age_kin + 1, "1900"]
head(
  check_mothers[
    check_mothers$age_kin >= 15,
    c("year", "age_focal", "kin", "age_kin",
      "death_over_living", "period_death_prob")
  ],
  8
)
```

The columns `death_over_living` and `period_death_prob` coincide because they describe the same interval: deaths of mothers during 1900 divided by mothers alive at the start of 1900. Living kin are therefore interpreted as a stock, while deaths are interpreted as a period flow.

Because deaths are flows over the interval after a stock is observed, the final stock in a time series has no following interval inside the model. Therefore, *deaths for the last input year are not available unless the model is projected one additional period*.

For the case of time-invariant models, all previous is also valid but only considering age of Focal, time is not included.
